A Reproducible Data Analysis Workflow with RMarkdown, Git, Make & Docker

### 2020-05-19 manuscript.Rmd
fix cross ref (again) closes #43

Figure \ref{fig:schematic} gives an overview of how the four components of our workflow interact to ensure computational reproducibility. Before we describe the four components in more detail, we begin with a minimal description of the roles of each component. In the remainder of this tutorial, we will further detail each of the four components of our workflow. The first component is version control. Version control manages changes to files (e.g., data and code) over time so that you can recall specific versions of files later or revert the entire project to a past state. Version control offers snapshots of your workflow at different time points identified by a unique identifier. The different frames in the z-axis of Figure  \ref{fig:schematic} illustrate the different versions of the project. How different parts of a analysis and a report of it relate and in what order they need to be executed is documented using dependency management. The arrows visualize dependencies, such as an analysis depending on the availability of a particular data file. Third, all computer code (such as a statistical analyses in R) is executed in a virtual environment that guarantuees exact reproduction of results independent of the host operating system, the locally installed R version, and installed package versions. Finally, dynamic document generation (also known as the literate programming paradigm) interweaves human-readable code and computed results (such as point estimates, p values, or confidence intervals) to eliminate inconsistency errors such as those arising from copy\&paste errors.

@@

The translation of computational results into a human-readable summary, for example into a technical report, a presentation, or a journal manuscript, is time-consuming and error-prone. Typical errors result from copy\&paste mistakes, erroneous rounding, or missed updates of the manuscript when the associated computer code and computed results have changed. In order to create not only fully reproducible results but also fully reproducible reports, we resort to the literate programming paradigm [@knuth1984literate], in which human-readable language and computer code are mixed to create dynamic documents whose order follows the logic of thought rather then the order of the computer. R Markdown is a simple markup language to create dynamic documents with embedded chunks of R code that can be exported to standard formats such as documents (docx, rtf, epub), presentations (ppt, html) or websites (html) using the `knitr` package [@xie2015; @xie2019]. Several packages extend the functionality of knitr. Of particular note are the `papaja` package [@papaja2018], which offers additional functions to enable American Psychological Association (APA) style document formatting, including a journal-style final typeset format, and the `stargazer` package [@stargazer2018], which provides journal-ready tables and reports of statistical models.  Figure \ref{fig:source-rmarkdown} illustrates R Markdown syntax using the `papaja` package and Figure \ref{fig:rendered-result} shows the resulting rendered document.

@@

```{r rendered-result, eval = TRUE, echo = FALSE,  fig.cap="Rendered result of the source code shown in Figure \ref{fig:source-rmarkdown}."}
knitr::include_graphics("R/rmarkdown.pdf")
```

@@

To keep track of all changes on your local computer, you only need to use `git add` and `git commit` or `git commit -a` to add and commit at the same time. Adding a file means to save its changes on the next commit. These commands need to be executed in the terminal, which you can access from within RStudio (`Shift + Alt + R`).  RStudio also offers a graphical user interface for Git. For most basic operations, this interface is convenient and sufficient (see Figure \ref{fig:gitpane}).

@@

`git push` or the green upward arrow in the Git pane (see Figure \ref{fig:gitpane}) uploads local updates. To download the remote Git repository on another computer, type into the terminal:

@@

The first line after the comment is the first (default) target called 'all', which depends on manuscript.pdf, which itself is a target. If Make is called without an argument, the first target is built. To create `manuscript.pdf` (the second target in the file), the file `manuscript.Rmd` needs to be rendered, which depends on `data/ìris_prepped.csv`. This dependency is itself a target (the third target in the file). To create `data/ìris_prepped.csv`, `R/prepare_data.R` and `data/iris.csv` are needed. If you type `make manuscript.pdf`, Make first checks whether the dependencies do exist and, if not, creates them. Here, if `data/ìris_prepped.csv` does not exist, Make creates it by executing the third target. Also, if one of the dependencies of a target is newer than the target itself, Make updates everything that directly or indirectly depends on the target. Here, if `data\iris.csv` is newer than `data\ìris_prepped.csv`, Make will attempt to recreate `data\ìris_prepped.csv` first and then `manuscript.pdf`. If there is a dependency missing, and there is no target to make it, Make stops with an error message. It is a convention to have the first target named `all`, which creates the entire project. Subsequently, the command `make` without any argument automatically creates everything possible in the project. The button `Build All` from within RStudio triggers this process (see Figure \ref{fig:buildpane-rstudio}).

@@

The `--rm` flag means that the state of the container after the command will have finished is not going to be saved. The `-it` flag tells Docker to run the command interactively, that is, to accept keyboard inputs and return outputs to the terminal. For example, this is the command to start an interactive R session inside a Docker image called `reproducible-research` (see Figure \ref{fig:docker-r-terminal} for a screenshot):

@@

Figure \ref{fig:rstudio-in-docker} shows a screenshot of Rstudio running inside Docker accessed from a local web browser.

>>>>

Figure \@ref(fig:schematic) gives an overview of how the four components of our workflow interact to ensure computational reproducibility. Before we describe the four components in more detail, we begin with a minimal description of the roles of each component. In the remainder of this tutorial, we will further detail each of the four components of our workflow. The first component is version control. Version control manages changes to files (e.g., data and code) over time so that you can recall specific versions of files later or revert the entire project to a past state. Version control offers snapshots of your workflow at different time points identified by a unique identifier. The different frames in the z-axis of Figure  \@ref(fig:schematic) illustrate the different versions of the project. How different parts of a analysis and a report of it relate and in what order they need to be executed is documented using dependency management. The arrows visualize dependencies, such as an analysis depending on the availability of a particular data file. Third, all computer code (such as a statistical analyses in R) is executed in a virtual environment that guarantuees exact reproduction of results independent of the host operating system, the locally installed R version, and installed package versions. Finally, dynamic document generation (also known as the literate programming paradigm) interweaves human-readable code and computed results (such as point estimates, p values, or confidence intervals) to eliminate inconsistency errors such as those arising from copy\&paste errors.

@@

The translation of computational results into a human-readable summary, for example into a technical report, a presentation, or a journal manuscript, is time-consuming and error-prone. Typical errors result from copy\&paste mistakes, erroneous rounding, or missed updates of the manuscript when the associated computer code and computed results have changed. In order to create not only fully reproducible results but also fully reproducible reports, we resort to the literate programming paradigm [@knuth1984literate], in which human-readable language and computer code are mixed to create dynamic documents whose order follows the logic of thought rather then the order of the computer. R Markdown is a simple markup language to create dynamic documents with embedded chunks of R code that can be exported to standard formats such as documents (docx, rtf, epub), presentations (ppt, html) or websites (html) using the `knitr` package [@xie2015; @xie2019]. Several packages extend the functionality of knitr. Of particular note are the `papaja` package [@papaja2018], which offers additional functions to enable American Psychological Association (APA) style document formatting, including a journal-style final typeset format, and the `stargazer` package [@stargazer2018], which provides journal-ready tables and reports of statistical models.  Figure \@ref(fig:source-rmarkdown) illustrates R Markdown syntax using the `papaja` package and Figure \@ref(fig:rendered-result) shows the resulting rendered document.

@@

```{r rendered-result, eval = TRUE, echo = FALSE,  fig.cap="Rendered result of the source code shown in Figure \\ref{fig:source-rmarkdown}."}
knitr::include_graphics("R/rmarkdown.pdf")
```

@@

To keep track of all changes on your local computer, you only need to use `git add` and `git commit` or `git commit -a` to add and commit at the same time. Adding a file means to save its changes on the next commit. These commands need to be executed in the terminal, which you can access from within RStudio (`Shift + Alt + R`).  RStudio also offers a graphical user interface for Git. For most basic operations, this interface is convenient and sufficient (see Figure \@ref(fig:gitpane)).

@@

`git push` or the green upward arrow in the Git pane (see Figure \@ref(fig:gitpane)) uploads local updates. To download the remote Git repository on another computer, type into the terminal:

@@

The first line after the comment is the first (default) target called 'all', which depends on manuscript.pdf, which itself is a target. If Make is called without an argument, the first target is built. To create `manuscript.pdf` (the second target in the file), the file `manuscript.Rmd` needs to be rendered, which depends on `data/ìris_prepped.csv`. This dependency is itself a target (the third target in the file). To create `data/ìris_prepped.csv`, `R/prepare_data.R` and `data/iris.csv` are needed. If you type `make manuscript.pdf`, Make first checks whether the dependencies do exist and, if not, creates them. Here, if `data/ìris_prepped.csv` does not exist, Make creates it by executing the third target. Also, if one of the dependencies of a target is newer than the target itself, Make updates everything that directly or indirectly depends on the target. Here, if `data\iris.csv` is newer than `data\ìris_prepped.csv`, Make will attempt to recreate `data\ìris_prepped.csv` first and then `manuscript.pdf`. If there is a dependency missing, and there is no target to make it, Make stops with an error message. It is a convention to have the first target named `all`, which creates the entire project. Subsequently, the command `make` without any argument automatically creates everything possible in the project. The button `Build All` from within RStudio triggers this process (see Figure \\@ref(fig:buildpane-rstudio)).

@@

The `--rm` flag means that the state of the container after the command will have finished is not going to be saved. The `-it` flag tells Docker to run the command interactively, that is, to accept keyboard inputs and return outputs to the terminal. For example, this is the command to start an interactive R session inside a Docker image called `reproducible-research` (see Figure \@ref(fig:docker-r-terminal) for a screenshot):

@@

Figure \@ref(fig:rstudio-in-docker) shows a screenshot of Rstudio running inside Docker accessed from a local web browser.
